---
title: "p8105_hw2_jw4348"
author: "Jingyu Wang"
output: github_document
date: "2023-10-04"
---

```{r, echo = FALSE, message = FALSE}
  library(tidyverse)
  library(readxl)
```

## Problem 1: 

- First, we clean the data in `pols-month.csv.`.

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
    )
```

```{r}
pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
```

- Then, we clean the data in `snp.csv`.

```{r clean_538_snp}
snp = 
  read_csv(
    "./data/fivethirtyeight_datasets/snp.csv",
    col_types = cols(date = col_date(format = "%m/%d/%y"))) |>
  separate(date, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    year = if_else(year > 2023, year - 100, year)) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, close) 
```

- Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

- Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

- Let's talk about the 538 datasets. The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r pols |> pull(year) |> min()` to `r pols |> pull(year) |> max()`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r snp |> pull(year) |> min()` to `r snp |> pull(year) |> max()`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r unemployment |> pull(year) |> min()` to `r unemployment |> pull(year) |> max()`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

## Problem 2: 

- First we read and clean the `Mr. Trash Wheel sheet`.

```{r}
MrTrashWheel_df = 
  readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |> 
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    type = "Mr Trash Wheel",
    year = as.double(year)
  )
 
MrTrashWheel_df
```

- Next we read and clean the `Professor Trash Wheel`.

```{r}
ProfessorTrashWheel_df = 
  readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |> 
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    type = "Professor Trash Wheel"
    )

ProfessorTrashWheel_df
```

- Then we read and clean the `Gwynnda Trash Wheel`.

```{r}
GwynndaTrashWheel_df = 
  readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet =   "Gwynnda Trash Wheel", range = "A2:L157") |> 
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    type = "Gwynnda Trash Wheel"
  )

GwynndaTrashWheel_df
```

- Now we merge the three datasets, and proce a single tidy dataset named `Trashtidy_df`.

```{r}
Trashtidy_df = 
  bind_rows(MrTrashWheel_df, ProfessorTrashWheel_df, GwynndaTrashWheel_df) |>
  janitor::clean_names() 

Trashtidy_df
```

- Blow is the description of dataset `Trashtidy_df`:
  - The number of observations in the resulting dataset `Trashtidy_df` is **`r nrow(Trashtidy_df)`**. 
  - There are **`r ncol(Trashtidy_df)`** variables.
  - The key variables in this dataset are **dumpster**, **month**, **year**, **date** and etc. 
  - All the variables are **`r colnames(Trashtidy_df)`**. 
  - The total weight of trash collected by Professor Trash Wheel was **`r sum(filter(Trashtidy_df,type == "Professor Trash Wheel") |> pull(weight_tons))`** tons.
  - The total number of cigarette butts collected by Gwynnda in July of 2021 was **`r sum(filter(Trashtidy_df,type == "Gwynnda Trash Wheel", month == "July", year == "2021") |> pull(cigarette_butts))`**.